Follow-up Samples

### 2017-09-15 submission/manuscript_R2.Rmd

Despite growing evidence that members of the gut microbiota can drive tumorigenesis, little is known about how treatment impacts the colonic community. This study tested the hypothesis that treatment for adenoma or carcinoma alters the abundance of bacterial populations associated with disease to those associated with a normal colon. We tested this hypothesis by sequencing the 16S rRNA genes in the feces of `r length(rownames(good_metaf))` individuals before and after treatment for adenoma (N = `r change_summary["thetayc", "adn_n"]`), advanced adenoma (N = `r change_summary["thetayc", "srn_n"]`), and carcinoma (N = `r change_summary["thetayc", "crc_n"]`). The structure of the microbiota in all three groups changed following treatment (P`r ifelse(change_summary["thetayc", "Pvalue_adnVcrc"]<0.001,"<0.001", format(change_summary["thetayc", "Pvalue_adnVcrc"], scientific = FALSE, digits = 2, nsmall = 2))`). To determine whether individuals who underwent treatment were more likely to have OTUs associated with normal colons we used a larger cohort that contained individuals with normal colons (N=XXX) and untreated individuals with adenomas (N=XXX), advanced adenomas (N=XXX), and carcinomas (N=XXX). Using these data, we built Random Forest models and measured the change in the positive probability of having one of the three diagnoses to assess whether the post-treatment samples received the same classification as the pre-treatment samples. The microbiota of patients diagnosed with carcinomas changed towards that of a normal colon following treatment (P<0.XXX). Those were initially diagnosed as having adenomas or advanced adenomas were no more likely to have a microbiota resembling that of a normal colon. These results further illuminate our understanding of the role of the gut microbiota in tumorigenesis and demonstrate the potential for using the gut microbiota as biomarkers for assessing risk and the effectiveness of treatment.

## Importance

Colorectal cancer is a worldwide health problem. Previous work in animal models and human populations has demonstrated that the gut microbiota has an important role in promoting tumorigenesis through a variety of mechanisms. Less well understood is how the gut microbiota changes following treatment for varying stages of tumorigenesis. We observed that following treatment of patients diagnosed with carcinoma, their gut microbiota is likely revert to a state associated with individuals who had normal colons whereas those diagnosed with adenomas and advanced adenomas were not. This study was a necessary step to ultimately investigating how the microbiota changes in those with carcinoma recurrence and predicting which patients are more likely to recur. Finding better strategies to stratify patients based on recurrence risk could provide a more effective means of planning therapies and monitoring patients after treatment.

>>>>

**Background.** Colorectal cancer is a worldwide health problem. Despite growing evidence that members of the gut microbiota can drive tumorigenesis, little is known about what happens to it after treatment for an adenoma or carcinoma. This study tested the hypothesis that treatment for adenoma or carcinoma alters the abundance of bacterial populations associated with disease to those associated with a normal colon. We tested this hypothesis by sequencing the 16S rRNA genes in the feces of `r length(rownames(good_metaf))` individuals before and after treatment for adenoma (N = `r change_summary["thetayc", "adn_n"]`), advanced adenoma (N = `r change_summary["thetayc", "srn_n"]`), and carcinoma (N = `r change_summary["thetayc", "crc_n"]`).

**Results.** There were small changes to the bacterial community associated with adenoma or advanced adenoma and large changes associated with carcinoma. The communities from patients with carcinomas changed significantly more than those with adenoma following treatment (P-value `r ifelse(change_summary["thetayc", "Pvalue_adnVcrc"]<0.001,"< 0.001", format(change_summary["thetayc", "Pvalue_adnVcrc"], scientific = FALSE, digits = 2, nsmall = 2))`). Although treatment was associated with intrapersonal changes, the change in the abundance of individual OTUs in response to treatment was not consistent within diagnosis groups (P-value `r ifelse(min(all_OTU_pair_wilcox$BH_corrected, na.rm = TRUE) < 0.05, invisible("<"), invisible(">"))` 0.05). Because the distribution of OTUs across patients and diagnosis groups was irregular, we used the Random Forest machine learning algorithm to identify groups of OTUs that could be used to classify pre and post-treatment samples for each of the diagnosis groups. Although the three models could differentiate between the pre and post-treatment samples, there was little overlap between the OTUs that were indicative of each treatment. To determine whether individuals who underwent treatment were more likely to have OTUs associated with normal colons we used a larger cohort that contained individuals with normal colons and those with adenomas, advanced adenomas, and carcinomas. We again built Random Forest models and measured the change in the positive probability of having one of the three diagnoses to assess whether the post-treatment samples received the same classification as the pre-treatment samples. Samples from patients who had carcinomas changed towards a microbial milieu that resembles the normal colon after treatment (P-value `r ifelse(all_probs_paired_wilcox["carcinoma", "BH_correction"] < 0.05, invisible("<"), invisible(">"))` 0.05). Finally, we were unable to detect any significant differences in the microbiota of individuals treated with surgery alone and those treated with chemotherapy or chemotherapy and radiation (P-value `r ifelse(min(chemo_rad_summary$bh, na.rm = TRUE) < 0.05, invisible("<"), invisible(">"))` 0.05).

**Conclusions.** By better understanding the response of the microbiota to treatment for adenomas and carcinomas, it is likely that biomarkers will eventually be validated that can be used to quantify the risk of recurrence and the likelihood of survival. Although it was difficult to identify significant differences between pre and post-treatment samples from patients with adenoma and advanced adenoma, this was not the case for carcinomas. Not only were there large changes in pre versus post-treatment samples for those with carcinoma, but these changes were towards a more normal microbiota.
